Nutritional Deficits and the Quantity-Quality Trade-off: Evidence from an Exogenous Fertility Shock in Low-Income Urban Settings in the Philippines

### 2025-07-07 docs/paper.Rmd
revise data and empirical strategy section
The revision details dataset construction, outcome definitions, covariates, treatment assignment, empirical approach, and robustness checks. Also updated the difference-in-differences specification and missing data handling. Removed outdated content and improved methodological transparency.

>>>>

\section{Data and Empirical Strategy}

This study constructs a harmonized, child-level analytic dataset using five rounds of the Philippine Demographic and Health Survey (DHS): 1993, 1998, 2003, 2008, and 2013. The DHS is a nationally representative, cross-sectional survey series conducted by the Philippine Statistics Authority in collaboration with ICF International. Each round follows a stratified two-stage cluster design and collects standardized information on child health, fertility, maternal characteristics, and household infrastructure. For each wave, I merge the child-level (KR), mother-level (IR), and household-level (HR) files using cluster, household, and maternal identifiers, ensuring that each observation corresponds to a live birth occurring within five years of the survey. The analytic sample is restricted to children aged 0–59 months to standardize developmental exposure windows and conform to the World Health Organization (WHO) reference growth standards.

The primary outcomes are height-for-age (HAZ) and weight-for-height (WHZ) z-scores, internationally recognized proxies for chronic and acute malnutrition, respectively. These anthropometric indicators are cleaned following DHS protocols: biologically implausible values are excluded, and remaining z-scores are rescaled from hundredths to continuous units. In addition to anthropometry, the child record includes a suite of health-related variables. These include indicators for whether the child has received basic vaccinations—such as BCG, DPT, polio, and measles—as well as whether the child experienced episodes of diarrhea, fever, or respiratory illness in the two weeks prior to the survey. These supplementary health measures serve as intermediate outcomes in secondary analyses to capture short-run morbidity and preventive health investment.

Maternal-level information is drawn from the individual recode file. Core covariates include maternal age, completed years of education, and total number of children ever born (parity), each of which captures socioeconomic status and fertility history. I additionally include maternal body mass index (BMI) and anemia status, which serve as both controls and potential mediators of child health. These indicators reflect the mother’s nutritional status and physiological capacity to support healthy fetal and postnatal development. In robustness checks, I examine whether declines in maternal nutritional status following the fertility shock account for observed changes in child anthropometry.

Household-level characteristics are used to control for environmental and infrastructural determinants of health. Specifically, I include household size, urban residence, wealth quintile, access to electricity, type of toilet facility, whether the toilet is shared, and source of drinking water. These variables proxy for household-level resources and exposure to disease environments. In extension analyses, I interact treatment status with these environmental indicators to assess whether the nutritional consequences of the fertility shock are magnified under infrastructural deprivation.

The treatment is defined by exposure to Executive Order No. 003, which was issued by the Manila City government in January 2000 and effectively suspended the public distribution of modern contraceptives. This policy change resulted in a localized increase in fertility that was not mirrored in surrounding municipalities within the National Capital Region (NCR). Children born in Manila City after the implementation of the policy are classified as treated, while children in other NCR cities serve as the control group. This spatially and temporally bounded intervention provides quasi-experimental variation that facilitates identification of causal effects.

The empirical approach adopts a difference-in-differences framework that compares trends in child nutritional outcomes between treated and control cities before and after the policy change. The identifying assumption is that, absent the policy, nutritional trajectories in Manila and non-Manila NCR cities would have evolved in parallel. This assumption is supported by visual inspection and pre-trend placebo tests using the 1993 and 1998 waves. The 1998 DHS serves as the primary pre-policy baseline, while the 2003 survey captures short-run effects. The 2008 and 2013 waves are used in robustness checks to assess medium- and long-run persistence of nutritional deficits. All specifications include city and birth-year fixed effects, and standard errors are clustered at the city-year level to account for within-cluster correlation.

The empirical strategy exploits the 2000 issuance of Executive Order No. 003, which effectively banned the distribution of modern contraceptives in public health centers across Manila City. This abrupt and localized withdrawal of contraceptive access resulted in a fertility increase concentrated within the city, but did not affect surrounding municipalities within the National Capital Region (NCR). Children born in Manila following the policy are considered exposed to a fertility shock, while children in other NCR cities over the same period serve as the control group. The analytical sample is restricted to children under five years old at the time of each survey round to ensure accurate measurement of early-life nutritional outcomes.

I estimate the following difference-in-differences specification:

\text{Nutrition}_{ict} = \alpha + \beta \cdot (\text{Post}_t \times \text{Manila}_c) + X_{ict}'\gamma + \lambda_c + \tau_t + \varepsilon_{ict}

In this equation, $\text{Nutrition}_{ict}$ refers to either height-for-age (HAZ) or weight-for-height (WHZ) for child $i$ in city $c$ born in year $t$. The indicator $\text{Post}_t$ equals 1 if the child was born after the implementation of Executive Order No. 003 in 2000, and $\text{Manila}c$ equals 1 if the child resides in Manila City. The interaction term captures the effect of being born in Manila after the contraceptive policy was enacted. The vector $X{ict}$ includes controls for child age in months, sex, birth order, maternal age, education, parity, body mass index (BMI), anemia status, household wealth quintile, household size, electricity access, water source, toilet type, shared sanitation status, and urban residence. City fixed effects $\lambda_c$ control for time-invariant characteristics at the locality level, while year fixed effects $\tau_t$ absorb aggregate time shocks. The coefficient of interest, $\beta$, represents the average treatment effect of the policy-induced fertility shock on child nutrition. Standard errors are clustered at the city-year level to account for potential autocorrelation within geographic units over time.

To support the internal validity of the design, I use Philippine Census microdata from 1990, 1995, 2000, 2007, and 2010 to confirm the magnitude and timing of the fertility shock. The 1995 and 2007 waves provide clean pre- and post-policy benchmarks, while the 2000 census—although temporally proximate to the policy—is retained for robustness checks that test the sensitivity of estimates to the exact onset of fertility changes. A placebo comparison between 1990 and 1995 shows no divergence in fertility trends between Manila and control cities prior to the policy, strengthening the credibility of the parallel trends assumption.

To supplement the DHS and Census data, I incorporate the Annual Poverty Indicators Survey (APIS), a nationally representative household survey that captures detailed information on food expenditure, maternal education, and living standards. Although the APIS lacks anthropometric data and has more limited geographic coverage, it enables exploratory analyses of treatment effect heterogeneity by income and education strata and helps triangulate changes in maternal resource constraints following the policy shock.

Finally, missing data are handled following standard practices in applied microeconometrics. Observations with missing or flagged anthropometric outcomes are excluded from the analysis. For other covariates with partial nonresponse—such as maternal BMI, anemia, or household infrastructure—I adopt the missing-indicator method: missing values are imputed with a constant, and a binary indicator is included in all regression models to control for the fact of missingness. This approach preserves sample size while allowing the data to inform whether nonresponse is systematically related to the outcomes of interest. As robustness checks, I estimate alternative specifications using multiple imputation and inverse-probability weighting to test the sensitivity of results to different missing data assumptions. All descriptive and inferential statistics are computed using DHS-provided sampling weights, and models account for the complex survey design by clustering standard errors at the primary sampling unit.

### 2025-07-07 docs/paper.tex
revise data and empirical strategy section
The revision details dataset construction, outcome definitions, covariates, treatment assignment, empirical approach, and robustness checks. Also updated the difference-in-differences specification and missing data handling. Removed outdated content and improved methodological transparency.

>>>>

\section{Data and Empirical Strategy}

This study constructs a harmonized, child-level analytic dataset using
five rounds of the Philippine Demographic and Health Survey (DHS): 1993,
1998, 2003, 2008, and 2013. The DHS is a nationally representative,
cross-sectional survey series conducted by the Philippine Statistics
Authority in collaboration with ICF International. Each round follows a
stratified two-stage cluster design and collects standardized
information on child health, fertility, maternal characteristics, and
household infrastructure. For each wave, I merge the child-level (KR),
mother-level (IR), and household-level (HR) files using cluster,
household, and maternal identifiers, ensuring that each observation
corresponds to a live birth occurring within five years of the survey.
The analytic sample is restricted to children aged 0--59 months to
standardize developmental exposure windows and conform to the World
Health Organization (WHO) reference growth standards.

The primary outcomes are height-for-age (HAZ) and weight-for-height
(WHZ) z-scores, internationally recognized proxies for chronic and acute
malnutrition, respectively. These anthropometric indicators are cleaned
following DHS protocols: biologically implausible values are excluded,
and remaining z-scores are rescaled from hundredths to continuous units.
In addition to anthropometry, the child record includes a suite of
health-related variables. These include indicators for whether the child
has received basic vaccinations---such as BCG, DPT, polio, and
measles---as well as whether the child experienced episodes of diarrhea,
fever, or respiratory illness in the two weeks prior to the survey.
These supplementary health measures serve as intermediate outcomes in
secondary analyses to capture short-run morbidity and preventive health
investment.

Maternal-level information is drawn from the individual recode file.
Core covariates include maternal age, completed years of education, and
total number of children ever born (parity), each of which captures
socioeconomic status and fertility history. I additionally include
maternal body mass index (BMI) and anemia status, which serve as both
controls and potential mediators of child health. These indicators
reflect the mother's nutritional status and physiological capacity to
support healthy fetal and postnatal development. In robustness checks, I
examine whether declines in maternal nutritional status following the
fertility shock account for observed changes in child anthropometry.

Household-level characteristics are used to control for environmental
and infrastructural determinants of health. Specifically, I include
household size, urban residence, wealth quintile, access to electricity,
type of toilet facility, whether the toilet is shared, and source of
drinking water. These variables proxy for household-level resources and
exposure to disease environments. In extension analyses, I interact
treatment status with these environmental indicators to assess whether
the nutritional consequences of the fertility shock are magnified under
infrastructural deprivation.

The treatment is defined by exposure to Executive Order No.~003, which
was issued by the Manila City government in January 2000 and effectively
suspended the public distribution of modern contraceptives. This policy
change resulted in a localized increase in fertility that was not
mirrored in surrounding municipalities within the National Capital
Region (NCR). Children born in Manila City after the implementation of
the policy are classified as treated, while children in other NCR cities
serve as the control group. This spatially and temporally bounded
intervention provides quasi-experimental variation that facilitates
identification of causal effects.

The empirical approach adopts a difference-in-differences framework that
compares trends in child nutritional outcomes between treated and
control cities before and after the policy change. The identifying
assumption is that, absent the policy, nutritional trajectories in
Manila and non-Manila NCR cities would have evolved in parallel. This
assumption is supported by visual inspection and pre-trend placebo tests
using the 1993 and 1998 waves. The 1998 DHS serves as the primary
pre-policy baseline, while the 2003 survey captures short-run effects.
The 2008 and 2013 waves are used in robustness checks to assess medium-
and long-run persistence of nutritional deficits. All specifications
include city and birth-year fixed effects, and standard errors are
clustered at the city-year level to account for within-cluster
correlation.

The empirical strategy exploits the 2000 issuance of Executive Order
No.~003, which effectively banned the distribution of modern
contraceptives in public health centers across Manila City. This abrupt
and localized withdrawal of contraceptive access resulted in a fertility
increase concentrated within the city, but did not affect surrounding
municipalities within the National Capital Region (NCR). Children born
in Manila following the policy are considered exposed to a fertility
shock, while children in other NCR cities over the same period serve as
the control group. The analytical sample is restricted to children under
five years old at the time of each survey round to ensure accurate
measurement of early-life nutritional outcomes.

I estimate the following difference-in-differences specification:

\text{Nutrition}\_\{ict\} = \alpha + \beta \cdot (\text{Post}\_t
\times \text{Manila}\emph{c) + X}\{ict\}'\gamma + \lambda\_c +
\tau\emph{t + \varepsilon}\{ict\}

In this equation, \(\text{Nutrition}_{ict}\) refers to either
height-for-age (HAZ) or weight-for-height (WHZ) for child \(i\) in city
\(c\) born in year \(t\). The indicator \(\text{Post}_t\) equals 1 if
the child was born after the implementation of Executive Order No.~003
in 2000, and \(\text{Manila}c\) equals 1 if the child resides in Manila
City. The interaction term captures the effect of being born in Manila
after the contraceptive policy was enacted. The vector \(X{ict}\)
includes controls for child age in months, sex, birth order, maternal
age, education, parity, body mass index (BMI), anemia status, household
wealth quintile, household size, electricity access, water source,
toilet type, shared sanitation status, and urban residence. City fixed
effects \(\lambda_c\) control for time-invariant characteristics at the
locality level, while year fixed effects \(\tau_t\) absorb aggregate
time shocks. The coefficient of interest, \(\beta\), represents the
average treatment effect of the policy-induced fertility shock on child
nutrition. Standard errors are clustered at the city-year level to
account for potential autocorrelation within geographic units over time.

To support the internal validity of the design, I use Philippine Census
microdata from 1990, 1995, 2000, 2007, and 2010 to confirm the magnitude
and timing of the fertility shock. The 1995 and 2007 waves provide clean
pre- and post-policy benchmarks, while the 2000 census---although
temporally proximate to the policy---is retained for robustness checks
that test the sensitivity of estimates to the exact onset of fertility
changes. A placebo comparison between 1990 and 1995 shows no divergence
in fertility trends between Manila and control cities prior to the
policy, strengthening the credibility of the parallel trends assumption.

To supplement the DHS and Census data, I incorporate the Annual Poverty
Indicators Survey (APIS), a nationally representative household survey
that captures detailed information on food expenditure, maternal
education, and living standards. Although the APIS lacks anthropometric
data and has more limited geographic coverage, it enables exploratory
analyses of treatment effect heterogeneity by income and education
strata and helps triangulate changes in maternal resource constraints
following the policy shock.

Finally, missing data are handled following standard practices in
applied microeconometrics. Observations with missing or flagged
anthropometric outcomes are excluded from the analysis. For other
covariates with partial nonresponse---such as maternal BMI, anemia, or
household infrastructure---I adopt the missing-indicator method: missing
values are imputed with a constant, and a binary indicator is included
in all regression models to control for the fact of missingness. This
approach preserves sample size while allowing the data to inform whether
nonresponse is systematically related to the outcomes of interest. As
robustness checks, I estimate alternative specifications using multiple
imputation and inverse-probability weighting to test the sensitivity of
results to different missing data assumptions. All descriptive and
inferential statistics are computed using DHS-provided sampling weights,
and models account for the complex survey design by clustering standard
errors at the primary sampling unit.
